---
title: "Greece COVID-19 Border Testing"
author: "Aidan Cooper"
date: "06/11/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(magrittr)
library(here)
library(ggtext)
library(mgcv)

theme_set(theme_classic())
# Set1: "#E41A1C" "#377EB8" "#4DAF4A" "#984EA3" "#FF7F00

# for plotting
width <- 8
height <- width * 9 / 16
```

## Background and Data

Research paper: [Efficient and targeted COVID-19 border testing via reinforcement learning](https://www.nature.com/articles/s41586-021-04014-z_reference.pdf)

Data sources:

 * [Greece border testing](https://github.com/kimondr/EVA_Public_Data)
 * [Our World in Data - COVID-19](https://github.com/owid/covid-19-data/tree/master/public/data)


Load data set.

```{r greece_data, message=FALSE}
df <- read_csv(here("data", "greece_eva.csv")) %>%
  separate(epoch_dates, sep = "/", into = c("epoch_start", "epoch_end")) %>%
  select(-c(epoch, epoch_end)) %>%
  mutate(
    epoch_start = as.Date(epoch_start),
    point_entry = str_split(point_entry, pattern="_") %>% sapply(extract, 2) %>% as.numeric()
  )
slice_sample(df, n=10)
```

There are 38 entry points, numbered from 1-41, with numbers 7, 32, and 35 not observed in the data.

```{r}
point_entry_nums <- sort(unique(df$point_entry))
paste(c(length(point_entry_nums), "entry points:", point_entry_nums), collapse=" ")
```

Create a filtered version containing only countries with at least 100,000 arrivals.

```{r}
# identify countries with at least 100,000 arrivals in Greece
countries <- df %>% 
  select(c(country, numSchedArrivals)) %>% 
  group_by(country) %>% 
  summarise(n_arrivals = sum(numSchedArrivals)) %>% 
  filter(n_arrivals > 100000) %>% 
  pull(country)
df_100k <- df %>% filter(country %in% countries)
paste(c(length(countries), "countries have at least 100,000 arrivals in Greece:", countries), collapse=" ")
```
The majority of arrivals from Bulgaria - Greece's neighbour - are at entry point number 1. The only other top 15 country with a notable proportion of arrivals at this entry point is Romania - presumably from tourists driving through Bulgaria. Similarly, neighbouring Albania also has most of its arrivals at entry point number 41, which barely features for any other top 15 country in the data set.

Entry point 17 is the only one with arrivals from all of the top 15 countries.

```{r}
df_heatmap <- df_100k %>%
  select(country, point_entry, numSchedArrivals) %>%
  group_by(country, point_entry) %>% 
  summarise(n_arrivals = sum(numSchedArrivals)) %>% 
  group_by(country) %>% mutate(n_arrivals = n_arrivals / sum(n_arrivals) * 100)
df_heatmap
f <- df_heatmap %>% ggplot(aes(x=country, y=point_entry, fill=n_arrivals)) + 
  geom_tile() +
  labs(
    title = "Entry points for each country",
    subtitle = paste(
      "Each country's percentage of arrivals at different Greek entry points varies considerably"
    ),
    x = "Origin country",
    y = "Entry point number",
    fill = "% arrivals",
    caption = "© Aidan Cooper 2021 | www.aidancooper.co.uk"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_markdown(),
    plot.caption = element_text(hjust = 0, size = 8,  colour = "#222222"),
    axis.title.y = element_text(vjust = 2, size = 12),
    panel.grid.major = element_line(),
    panel.grid.minor = element_line()
  )
ggsave("../figures/entry_points_heatmap.png", width = width, height = height)
f
```

Entry points

```{r}
df_entry <- df_100k %>%
  select(-c(country, epoch_start)) %>%
  group_by(point_entry) %>%
  summarise(across(everything(), list(sum))) %>%
  mutate(
    tests_per_1000arrival = numTestsPerformed_1 / numSchedArrivals_1 * 1000,
    positives_per_1000test = numPositivesFound_1 / numTestsPerformed_1 * 1000,
    n_arrivals = numSchedArrivals_1,
    n_tests = numTestsPerformed_1,
  ) %>%
  select(c("point_entry", "n_arrivals", "n_tests", "tests_per_1000arrival", "positives_per_1000test")) %>%
  arrange(desc(n_arrivals))
top3_percent <- sum(df_entry$n_arrivals[1:3]) / sum(df_entry$n_arrivals) * 100

f <- df_entry %>% ggplot(aes(x=point_entry, y=n_arrivals)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Number of arrivals at each entry point",
    subtitle = paste0(
      "The three busiest entry points account for ",
      round(top3_percent, 0),
      "% of all arrivals"
    ),
    x = "Number of arrivals",
    y = "Entry point number",
    caption = "© Aidan Cooper 2021 | www.aidancooper.co.uk"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_markdown(),
    plot.caption = element_text(hjust = 0, size = 8,  colour = "#222222"),
    axis.title.y = element_text(vjust = 2, size = 12),
    panel.grid.major = element_line(),
    panel.grid.minor = element_line()
  )
ggsave("../figures/entry_points_n_arrivals.png", width = width, height = height)
f
```


The number of tests conducted on each day of the week is consistent across all days, but the number of arrivals is considerably higher on Thu-Sun. This means that the rate of testing is higher for Thu-Sun than for Mon-Wed.

```{r}
df_tests <- df %>%
  select(-c(country, point_entry)) %>%
  group_by(epoch_start) %>%
  summarise(across(everything(), list(sum))) %>%
  mutate(
    tests_per_1000arrival = numTestsPerformed_1 / numSchedArrivals_1 * 1000,
    positives_per_1000test = numPositivesFound_1 / numTestsPerformed_1 * 1000,
    n_arrivals = numSchedArrivals_1,
    n_tests = numTestsPerformed_1,
    weekday = lubridate::wday(epoch_start, week_start = 1) < 4
  ) %>%
  select(c("epoch_start", "n_arrivals", "n_tests", "tests_per_1000arrival", "positives_per_1000test", "weekday")) %>%
  arrange(epoch_start)
mod1 <- gam(tests_per_1000arrival ~ s(as.numeric(epoch_start), bs = "cs"), data = df_tests)
df_tests <- df_tests %>%
  mutate(
    gam = mod1$fitted.values,
    gam_percent_diff = (tests_per_1000arrival - gam) / tests_per_1000arrival * 100
  )
df_dow <- df_tests %>%
  ungroup() %>%
  mutate(day_of_week = lubridate::wday(epoch_start, label = TRUE, week_start = 1)) %>% 
  select(c("day_of_week", "gam_percent_diff", "n_arrivals", "n_tests")) %>%
  group_by(day_of_week) %>%
  summarise(
    mean_diff = mean(gam_percent_diff),
    total_arrivals = sum(n_arrivals),
    total_tests = sum(n_tests)
    )
df_dow
```
Plot the _tests per 1,000 arrivals_ throughout the study period.

```{r}
f <- ggplot(df_tests, aes(x = epoch_start, y = tests_per_1000arrival)) +
  geom_line() +
  geom_point(
    mapping = aes(colour = weekday)
  ) +
  geom_line(aes(y = gam), colour = "grey", linetype = "dashed", size = 1) +
  labs(
    title = "COVID-19 testing per 1,000 arrivals",
    subtitle = paste(
      "Testing increased significantly over time, with higher testing rates on ",
      "**<span style='color:#377EB8;'>Mon-Wed</span>** than ",
      "**<span style='color:#E41A1C;'>Thu-Sun</span>**"
    ),
    x = "",
    y = "Tests performed per 1,000 arrivals",
    caption = "© Aidan Cooper 2021 | www.aidancooper.co.uk"
  ) +
  guides(colour = "none") +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_markdown(),
    plot.caption = element_text(hjust = 0, size = 8,  colour = "#222222"),
    axis.title.y = element_text(vjust = 2, size = 12),
    panel.grid.major = element_line(),
    panel.grid.minor = element_line()
  ) +
  scale_color_brewer(palette = "Set1") +
  scale_x_date(date_breaks = "1 month", limits = as.Date(c("2020-08-01", "2020-11-01")))
ggsave("../figures/testing_over_time.png", width = width, height = height)
f
```


Our World in Data COVID-19 dataset.

```{r covid_data, message=FALSE}
df_covid <- read_csv(here("data", "owid-covid-data.csv")) %>%
  select(c(
    iso_code,
    location,
    date,
    total_cases_per_million,
    new_cases_smoothed_per_million,
    total_deaths_per_million,
    new_deaths_smoothed_per_million,
    weekly_icu_admissions_per_million,
    weekly_hosp_admissions_per_million,
    new_tests_smoothed_per_thousand,
    positive_rate,
    tests_per_case,
    total_vaccinations_per_hundred
  ))
```
